---
title: "Producing Figure 4 for vaccination paper"
author: "Samik Datta, Giorgia Vattiato and Michael J. Plank"
editor: visual
date: "2023-12-06"
format: 
  html:
    embed-resources: true
    page-layout: full
    code-fold: true
    toc: true
    toc-location: left
    toc-title: Table of contents
    number-sections: true
    fig-width: 12
    fig-height: 10
knitr: 
  opts_chunk: 
    collapse: true
    echo: false
    warning: false
    message: false
---

### For further information, please contact

-   Samik Datta [(samik.datta\@niwa.co.nz)](mailto:samik.datta@niwa.co.nz)
-   Giorgia Vattiato [(vattiatogi@landcareresearch.co.nz)](mailto:vattiatogi@landcareresearch.co.nz)
-   Michael Plank [(michael.plank\@canterbury.ac.nz)](mailto:michael.plank@canterbury.ac.nz)

## Acknowledgements

This work represents the combined efforts of a large number of individuals and organisations, past and present, working within Covid-19 Modelling Aotearoa. Further information can be found [here](https://www.covid19modelling.ac.nz/about/).

## ODE model summary

The ODE model is based on numerous parameter assumptions: some of these values are fixed, others are fitted using a naive ABC (approximate Bayesian computation) method (see Lustig et al., 2023 for details).

Using this method, the model picks random combinations of parameter values from their 'prior' distributions (results presented here result from 15,000 independent draws from each prior distribution), then outputs a posterior distribution of accepted values. The accepted values are those resulting in the 1% best fitting trajectories (i.e. those that give the smallest value for a distance function that captures the difference between the infection/case/hospitalisation/etc. values of a simulated trajectory and those from observed data), and are then used to plot a 'best fit' line and a 95% confidence interval.

It is worth noting that multiple different combinations of parameter choices can result in simulations that give an equally good fit to empirical data, including combinations of parameters that may conflict with other parameter combinations. It is therefore important to consider modelling results as an ensemble of plausible trajectories, subject to the constraints and assumptions of the model.

We note that, in the latest model fit, we have adjusted our infection hospitalisation rates by age to match more closely with the observed hospitalisations.

### Data used in model fit

We fit our model to the following data:

-   daily infections in a routinely tested cohorted of border workers (up to July 2022),
-   aggregate and age-split daily reported cases, aggregate daily hospital admissions and aggregate daily deaths (up to 13th August 2023), and
-   vaccination data (up to 6th June 2023).

### Data plotted

The data points shown on plots include:

-   daily infections in a routinely tested cohorted of border workers up to July 2022,
-   aggregate and age-split daily reported cases, aggregate daily hospital admissions and aggregate daily deaths (up to 13th August 2023), and
-   aggregated hospital occupancy (up to 16th August 2023).

All data are plotted using a 7-day rolling average for smoothing.

### Scenarios simulated

The vaccination scenarios simulated here include:

1.  Baseline: vaccination rates as occurred in New Zealand. 2: No vaccination: No vaccination carried out. 3: No vaccination (under 60s): Removing vaccination for those aged under 60.
2.  90% vaccination: reducing vaccination rates to 90% of what they were in reality.
3.  Reduced coverage in older ages:
4.  Maori uptake rates: assuming vaccination occurred at the per-capita rate that it did for the Maori population (split into 5-year age groups).
5.  European / Other vaccination: assuming vaccination occurred at the per-capita rate that it did for the European / Other population (split into 5-year age groups).

All runs are performed from 5th March 2023 until 30th June 2023.ases, aggregate daily hospital admissions and aggregate daily deaths (up to 16th April 2023), and
-   vaccination data (up to 24th April 2023).

```{r}

#| label: read_in_data


# libraries needed
library(tidyverse)
library(readxl)
library(grid)
options(scipen = 999) # avoid scientific notation

# Read in data ####
filename_data = '../results/results_for_vaccination_scenarios_paper/compare_scenarios_20231016.xlsx'
filename_age = '../results/results_for_vaccination_scenarios_paper/compare_scenarios_20231016_age_split.xlsx' # filename pointers

# Default values
text_size_aggregate = 16
text_size_age_split = 16
col_palette = 'Set1'
data_col = 'black'
plot_min_date = '2022-01-01'
plot_max_date = '2023-06-30'

scen_order = c("Baseline", "No vaccine", "No AVs", "No vaccine or AVs", "No vaccine in U60s", 
               "10% drop in rates", "20-25-year-old rates", "MÄori rates", "Euro/other rates")


# Load plotting function

plot_compare_multiple = function(metric = c('infections', 'cases', 'admissions', 'deaths'), plot_pick = 'diff', min_date = NULL, max_date = NULL, actual_dates = T) {
  
  if (is.null(min_date)) {
    min_date = min(covid_tibble$date)
  } else {
    min_date = as.Date(min_date)
  }
  
  if (is.null(max_date)) {
    max_date = max(covid_tibble$date)
  } else {
    max_date = as.Date(max_date)
  }
  
  if (!all(metric %in% c('infections', 'cases', 'admissions', 'deaths'))) {
    stop('Metric needs to include either infections, cases, admissions or deaths.')
  }
  
  if (!plot_pick %in% c('median', 'diff', 'diffPerc')) {
    stop('plot_pick needs to be either median, diff or diffPerc.')
  }
  
  graph_titles = tibble(output = paste0(metric, '_', plot_pick),
                        titles = metric) %>%
    mutate(titles = factor(titles, levels = titles))
  
  test = covid_tibble %>% left_join(graph_titles, by = 'output') %>%
    filter(date >= min_date, date <= max_date, age == 'total', 
           scenario != 'Baseline',
           titles %in% metric) %>%
    droplevels()
  
  if (plot_pick == 'median') {
    y_title = 'Numbers'
    overall_title = 'Total numbers over time'
  } else if (plot_pick == 'diff') {
    y_title = 'Difference in numbers'
    overall_title = 'Difference in numbers from baseline'
  } else if (plot_pick == 'diffPerc') {
    y_title = '% difference'
    overall_title = '% Difference in numbers from baseline'
  }
  
  if (actual_dates == T) {
    
    ggplot(test, aes(x = date, colour = scenario, fill = scenario)) +
      geom_ribbon(aes(ymin = lower, ymax = upper), linetype = 0, alpha = 0.2) +
      geom_line(aes(y = value), linewidth = 1) +
      facet_wrap(~titles, scales = 'free_y', nrow = 2) +
      labs(x = 'Date', y = y_title, title = overall_title, colour = 'Scenario', fill = 'Scenario') +
      scale_fill_brewer(palette = 'Set1') +
      scale_color_brewer(palette = 'Set1') +
      theme_bw(base_size = 20) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
      scale_x_date(breaks = '3 months')
    
  } else {
    
    ggplot(test, aes(x = date_zero, colour = scenario, fill = scenario)) +
      geom_ribbon(aes(ymin = lower, ymax = upper), linetype = 0, alpha = 0.2) +
      geom_line(aes(y = value), linewidth = 1) +
      facet_wrap(~titles, scales = 'free_y', nrow = 2) +
      labs(x = 'Days since policy change', y = y_title, title = overall_title, colour = 'Scenario', fill = 'Scenario') +
      scale_fill_brewer(palette = 'Set1') +
      scale_color_brewer(palette = 'Set1') +
      theme_bw(base_size = 20) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
      scale_x_date(breaks = '3 months')
    
  }
  
}


# Read in data ####
all_data = read_excel(filename_data) # a bit weird due to cell arrays becoming columns
sheets = excel_sheets(filename_age) # get sheet names from age spreadsheet
age_data <- lapply(sheets, function(X) read_excel(filename_age, sheet = X)) # read in age split as list of tibbles
for (j in 1:length(age_data)) {
  names(age_data[[j]]) <- str_replace_all(names(age_data[[j]]), " ", "0")
}

# Specify dates for all columns - use date string from age split data
time_range = as.Date(age_data[[1]]$dates)

## Create table for aggregate variables ####

# Master table with no rows

total_tibble = tibble(date = Sys.Date(), output = character(), scenario = character(), age = character(), 
                      value = numeric(), lower = numeric(), upper = numeric())

# Setting up scenarios
all_scenarios = unique(as.character(all_data$scenario))

# Setting up outputs
all_outputs = c('infections', 'cases', 'admissions', 'deaths', 'occupancy')

# Setting up variables within each output
all_variables = c('median', 'diff', 'diffPerc')

# Do rest in loop

# Go through output - infections, cases, admissions, deaths

for (output_select in all_outputs) {
  
  # Go through variables - numbers, median, lower, upper
  
  for (variable_select in all_variables) {
    
    # go through scenarios
    
    for (scenario_pick in all_scenarios) {
      
      test = all_data %>% filter(scenario == scenario_pick) %>% 
        select(contains(paste0(output_select, variable_select, '_'))) # pull out columns
      dummy = tibble(date = time_range, output = paste0(output_select, '_', variable_select),
                     scenario = scenario_pick, age = 'total', value = as.numeric(test), lower = NA, upper = NA)
      
      if (variable_select == 'median') { # for best fit and median, no variable_select in string
        dummy = dummy %>% mutate(lower = as.numeric(all_data %>% filter(scenario == scenario_pick) %>% 
                                                      select(contains(paste0(output_select, 'lower', '_')))),
                                 upper = as.numeric(all_data %>% filter(scenario == scenario_pick) %>% 
                                                      select(contains(paste0(output_select, 'upper', '_')))))
      } else if (str_detect(variable_select, 'diff')) { # for diff and diffPerc, use variable select in name
        dummy = dummy %>% mutate(lower = as.numeric(all_data %>% filter(scenario == scenario_pick) %>% 
                                                      select(contains(paste0(output_select, variable_select, 'lower', '_')))),
                                 upper = as.numeric(all_data %>% filter(scenario == scenario_pick) %>% 
                                                      select(contains(paste0(output_select, variable_select, 'upper', '_')))))
      }
      
      total_tibble = total_tibble %>% add_row(dummy) # whack onto master tibble
      
    }
  }
}

## Now do for age split ####

# Master table with no rows

age_tibble = tibble(date = Sys.Date(), output = character(), scenario = character(), age = character(), 
                    value = numeric(), lower = numeric(), upper = numeric())


# Age groups
age_groups = c('0-9', '10-19', '20-29', '30-39', '40-49', '50-59', '60-69', '70+')

# Do rest in loop

sheet_number = 0 # start sheet counter

# Go through scenarios

for (scenario_pick in 1:length(all_scenarios)) {
  
  sheet_number = sheet_number + 1 # increase sheet counter
  
  # Go through output - infections, cases, admissions, deaths
  
  for (output_select in all_outputs) {
    
    # Go through variables - numbers, median, lower, upper
    
    for (variable_select in all_variables) {
      
      # Go through age groups, whack into 10-year groups (join 1:2m, 3:4, etc.)
      
      for (ages in 1:length(age_groups)) {
        
        dummy = tibble(date = time_range, output = paste0(output_select, '_', variable_select),
                       scenario = all_scenarios[scenario_pick], 
                       age = age_groups[ages], 
                       value = age_data[[sheet_number]] %>% 
                         pull(contains(paste0(output_select, variable_select, '_', ages))),
                       lower = NA, upper = NA) # pull out columns
        
        if (variable_select == 'median') { # for best fit and median, no variable_select in string
          dummy = dummy %>% mutate(lower = age_data[[scenario_pick]] %>% 
                                     pull(contains(paste0(output_select, 'lower_', ages))), # pull out columns
                                   upper = age_data[[scenario_pick]] %>% 
                                     pull(contains(paste0(output_select, 'upper_', ages))))
        } else if (str_detect(variable_select, 'diff')) { # for diff and diffPerc, use variable select in name
          dummy = dummy %>% mutate(lower = age_data[[scenario_pick]] %>% 
                                     pull(contains(paste0(output_select, variable_select, 'lower_', ages))), # pull out columns
                                   upper = age_data[[scenario_pick]] %>% 
                                     pull(contains(paste0(output_select, variable_select, 'upper_', ages))))
        }
        
        age_tibble = age_tibble %>% add_row(dummy) # whack onto master tibble
        
      }
    }
  }
}


## Now join tibbles and do admin for plots

covid_tibble = bind_rows(total_tibble, age_tibble) %>%
  replace_na(list(value = 0, lower = 0, upper = 0)) %>%
  left_join(tibble(scenario = all_scenarios, scen_order = factor(scen_order, levels = scen_order[c(1, 3, 2, 4, 6, 5, 7, 9, 8)])), by = 'scenario') %>%
  select(-scenario) %>%
  rename(scenario = scen_order) %>% # replace scenario number by name
  mutate(date_zero = as.numeric(date - min(date))) # days since policy change

```

```{r}

#| label: fig-admissions-and-deaths

plot_compare_multiple(metric = c('infections', 'admissions', 'deaths'), min_date = plot_min_date, max_date = plot_max_date) + 
  theme(legend.position = c(0.9, -0.1), legend.justification = c(1, 0))



```
